Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-12 manuscript.Rmd
First draft of intro and start Methods

---
title: A GIS based workflow for Ecosystem Condition Accounting - an example using spatially biased indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliation: Norwegian Institute for Nature Research
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - SEEA EA
  - alien species
  - ...
journal: "Ecological Indicators"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

# Introduction
SEEA EA principles

The need for local field data

Low data availability

Fast turn around with automated workflows

4 types of indicator

@@

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and  
a deep learning model.
This model had a XX% precision when tested against independent field data.

The reference state defining ecosystems in optimal or pristine condition is defined as one with little or no human influence, with a climate as in the period 1961-1990 and a native species similar as today (i.ei no mammoths allowed).

We then identified two _ecosystem condition characteristics_ that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition:

@@

To describe these two concepts we use five condition variables from the same data set (ref).
The five variables are:

1. 7FK (prevalence of alien species)
1. 7SE (tracks from large vehicles)
1. 7TK (human caused abration or abration-caused erosion)
1. PRTK (same as 7TK, but recorded at a higher resolution)
1. PRSL (same as 7SE, but recorded at a higher resolution)

Varaibles 1-3 are defined in Halvorsen and Bratli 2019.
Variables 4 and 5 are defined in ...

@@

Validation/sensitivity analyses (varying HIA resolution and minimum _n_?)

@@

# Results
Data availability/coverage

@@

# Credit authorship contribution statement {-}

# Declaration of Competing Interest {-}

# Acknowledgements {-}

# Data availability {-}

# References {-}

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliation: Norwegian Institute for Nature Research
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.
  It consists of two paragraphs.
keywords: 
  - SEEA EA
  - alien species
  - ...
journal: "Ecological Indicators"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
editor_options: 
  markdown: 
    wrap: sentence
---

# Introduction

Some general text about the decline in natural capital?

Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are variables and indicators.
These are parameters chosen to reflect the sentral condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 mean that the variable is the worst possible state.
The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthorpogenic disturbance.

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities [meaning areas] about which information is sought and about which statistics are ultimately compiled (REF). This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

a.  Using wall-to-wall data (e.g. remotely sensed data)
b.  Predict values using a model (e.g. by accounting for environmental variation)
c.  Simple projection of some best estimate, typically a central tendency from area representative data

The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often or not a pragmatic and opportunistic one which is unlikely to reflect the full scope of the ecosystem condition characteristics.
Note that the same is true for thematic biases.
For example reflected in the scarcity of data included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
Also having data from only one or a subset of nature types inside what is defined as the ecosystem in the assessment, is a typical thematic bias in ecosystem accounting.
However, in this paper we chose to focus on spatial bias.

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option b in the list above).
Models can describe the general associations between the spatially sampled data and the context (e.g. the set of environmental variables) where it was samples, and use these arelationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can be very reliable and make good indicators.
This is especially true when the ecosystem assets are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent lvel of uncertainty from such models becomes unacceptable.

In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across four neighboring municipalities in Norway.
The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.

@@

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and\
a deep learning model.
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire (Bakkestuen REF).
Mires are ecologically and socially important in Norway simply do to its large extent, and especially due to its role in climate mitigation.
Examples - how much carbon - (REFS) There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
This study builds on the work in that report.

We started by defining the reference state as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today (i.e no mammoths allowed).
We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition:

@@

To describe these two concepts we use five condition variables from the same data set (REF).
The five variables are:

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference  |
|------------|-------------|------------------|------------------|-------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | What fraction of the species composition is made up from aline species                                         |            |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          |            |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 1 to 8.                                        | Same as 7SE, but recorded at a higher resolution                                                               |            |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks |            |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 1 to 8.                                        | Same as 7TK, but recorded at a higher resolution                                                               |            |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to xxx                                                     | ...                                                                                                            | Bakkestuen |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   |                                                                                                                |            |

: Variables used in this study

@@

Validation/sensitivity analyses (varying HIA resolution and minimum *n*?)

@@

# Results

Data availability/coverage

@@

# Credit authorship contribution statement {.unnumbered}

# Declaration of Competing Interest {.unnumbered}

# Acknowledgements {.unnumbered}

# Data availability {.unnumbered}

# References {.unnumbered}
